Ensemble Transport Adaptive Importance Sampling

### 2019-01-14 response_JUQ_2.tex

\usepackage[colorlinks]{hyperref}

\newcommand{\comment}[2]{\vspace{0.6cm}{\bf Comment:} {\it #1.}

\vspace{0.3cm}{\bf Answer:} #2}

\begin{document}
\title{Response to Referee Comments: Parallel Adaptive Importance Sampling}
\maketitle
First of all, we would like to thank the referees and the associate editor for
their time in reading and commenting on our manuscript. We thank all of you and Professor Higdon for your patience in waiting for these delayed corrections. In this document
we will address each of the points raised, and describe any relevant
changes that have been made to the paper. We have made several substantial changes, including the renaming of the algorithm and paper from ``parallel adaptive importance sampling'' (PAIS) to ``ensemble transport adaptive importance sampling'' (ETAIS). We have also replaced two of the numerical examples. In particular we have incorporated a new bimodal example with differing scales in the two modes, and for which we have tested the ETAIS with three different resampling algorithms. We feel that this, in particular, makes the paper stronger, and we hope that the referees and editors will now agree that the paper is ready for publication.

@@

\comment{Also, the comparison between serial PAIS and the naive parallel MCMC is strange if there is no parallelization.}{Most of the numerics compares our methodology to Metropolis-Hastings, based on the number of likelihood evaluations required, rather than runtime, with the argument being based on the idea that this approach is most suitable for targets with expensive likelihood evaluations. Since the only difference between implementations in parallel or serial of our method will be in CPU runtime (due to extra overheads due to communications between processors), it is still valid for us to compare to a naively parallelised Metropolis-Hastings approach, since our choice of serial/parallel implementation makes no difference to these results. For the final example where we consider runtime, a sentence has been added to state that our results will be slightly overoptimistic based on the lack of communication overheads.}

\comment{It will be interesting that the authors show the PAIS using AMS with the examples. They show the difference between the AMS and the ETPF but not within the PAIS.}{We assume that the referee meant the AMR? As the AE asked for a new example to replace the second example, we took this as an opportunity to compare ETAIS with the ETPF, AMR and bootstrap resamplers, which is now presented in the new first example. The numerics demonstrate that superior resamplers lead to more stable and accurate importance sampling, with ETPF algorithm narrowly outperforming the AMR algorithm, with the bootstrap version following up in the rear. This helps support our claim that state-of-the-art resampling methods can lead to more efficient AIS algorithms. Note that we have changed the name of the AMR to the multinomial transform (MT).}

@@

\begin{figure}[htp]
\includegraphics[width=0.3\textwidth]{figures/xhist}
\includegraphics[width=0.3\textwidth]{figures/yhist}
\includegraphics[width=0.3\textwidth]{figures/zhist}
\caption{Histograms from 1e7 Monte Carlo runs of the push forward of the new prior for the Lorenz '63 problem}
\end{figure}

The more dimensions and posterior complexity that exists in a problem, the bigger the ensemble will need to be to have a stable ETAIS implementation. After a certain point, due to the extra overheads with a large ensemble due to for example the resampling, ETAIS will no longer be competitive. This said, in our most recent work, arbitrarily complex posteriors can be efficiently sampled using ETAIS if used in conjunction with a transport map to simplify the target. This is the subject of our next paper which we will be submitting imminently. We have added a paragraph to this end to the discussion.}

  \comment{8) Do not use an explicit Euler solver, use any higher order, off the shelve solver.}{In this example, we use explicit Euler applied to Lorenz '63 as a simple way to generate a discrete chaotic map so that the inverse problem is challenging. This is an identical twin experiment assuming that the discrete system is the perfect one from which the data is obtained (following the presentation of the textbook of Reich and Cotter), so changing to another integration scheme doesn't really change anything.}

  % {Many of the off-the-shelf higher order solvers are just as inappropriate to use for the approximation of the solution of the Lorenz '63 equations, since they are chaotic, and as such \emph{any} numerical error is quickly amplified through the dynamics. Moreover, the focus of this paper is not on these applications, but in the statistical algorithm. In this example, we are simply looking for a non-linear mapping such that we can demonstrate the speed-up that is gained from using ETAIS as opposed to standard M-H sampling, and we believe that a simple Euler approximation provides such a mapping.}

  \section{Other changes}
  We have also changed the name of the approximate multinomial resampler (AMR) to multinomial transformation (MT).

\end{document}

### 2019-01-14 ETAIS.tex
Removed AMR from figures
Final version complete

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Ensemble Transport Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. {\red As applications of these methods increase in size and
  complexity, the need for efficient methods increases. In this
paper, we present a particle ensemble algorithm.} At each iteration, an
  importance sampling proposal distribution is formed using an
  ensemble of particles. A stratified sample is taken from this
  distribution and weighted under the posterior, a state-of-the-art
  ensemble transport resampling method is then used to create an evenly weighted sample
  ready for the next iteration. We demonstrate that this ensemble transport
  adaptive importance sampling (ETAIS) method outperforms MCMC methods
  {\red with equivalent proposal distributions} for low dimensional problems,
  and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, {\red multinomial transformation (MT),} which while not
  as accurate as the ensemble transport resampler, is substantially less costly for large
  ensemble sizes, and can then be used in conjunction with ETAIS for
  complex problems. We also focus on how algorithmic parameters
  regarding the mixture proposal can be quickly tuned to optimise
  performance. In particular, we demonstrate this methodology's
  superior sampling for multimodal problems, such as those arising
  from inference for mixture models, and for problems with expensive
  likelihoods requiring the solution of a differential equation, {\red
    for
  which speed-ups of orders of magnitude are demonstrated. Likelihood
  evaluations of the ensemble could be computed in a distributed
  manner, suggesting that this methodology could be a good candidate
  for parallel Bayesian computations.}
\end{abstract}
\begin{keywords}MCMC, importance sampling, Bayesian, inverse problems,
  {\red ensemble, resampling}.
\end{keywords}
\section{Introduction}
Having first been developed in the early 1970s\cite{hastings1970monte}, Markov chain Monte Carlo (MCMC) methods have been of increasing
importance and interest in the last 20 years or so. They allow us to
sample from complex probability distributions which we would not be
able to sample from directly. In particular, these methods have
revolutionised the way in which inverse problems can be tackled,
allowing full posterior sampling when using a Bayesian framework.

@@

Then, according to Bayes' theorem, the posterior density that we
are interested in characterising is given, up to a constant of proportionality, by
\begin{equation*}
	\pi_Y \propto \exp \left ( -\Phi(x^*) \right )\pi_0(x^*),
\end{equation*}
where $\pi_0$ is the prior density on the quantity of interest $x^*$.

@@

\subsection{Particle filters and resamplers}\label{sec:filters}
In this subsection we briefly review particle filters, since the
development of the resampler that we incorporate into the ETAIS is
motivated by this area.

Particle filters are a class of Monte Carlo algorithms designed to
solve the filtering problem. That is, to find the best estimate of the
true state of a system when given only noisy observations of the
system. The solution of this problem has been of importance since the
middle of the 20th century in fields such as molecular biology,
computational physics and signal processing. In recent years the data
assimilation community has contributed several efficient particle
filters, including the ensemble Kalman filter
(EnKF)~\cite{evensen1994sequential} and the ensemble transform
particle filter (ETPF)~\cite{reich2013nonparametric}.

@@

In this work we use the resampler used within these particle filters. We
approximate the posterior {\red density $\pi$,} with a sample
of weighted particles, $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
the weights would be found by incorporating new observed data whereas
here we simply use importance weights. These weighted states can then be resampled, using the ETPF or otherwise,
into a {\red set of new equally-weighted states, such that
\begin{equation}\label{eqn:resampler}
	\sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
	 \frac{1}{M} \delta_{x_i}(\cdot).
\end{equation}}

The resampler used within particle filters such as the ETPF are well suited to this problem
since it is easy to introduce conditions in the resampling to ensure
you obtain the behaviour you require. One downside is that the
required ensemble size increases quickly with dimension, making it
difficult to use in high-dimensional problems.

@@

Initially we
will use the ETPF resampler algorithm\cite{reich2013nonparametric}, although we
will suggest an alternative strategy in Section \ref{sec:MT} where a
large ensemble is required, and for which the ETPF may become more expensive. The
resampling algorithm gives us a set of evenly weighted samples which
represents the target distribution well, which we can use to iterate
the process again. The algorithm is summarised in Algorithm
\ref{alg:ETAIS}.

{\red The choice of resampling algorithm is important since its
  output is used to formulate the proposal distribution for the next
  iteration. A basic resampler may be cheap to implement, but the
  resulting sample, and in turn the next iteration's proposal distribution, may not be as representative of the target
  distribution. This leads to higher variances of the weights in the
  importance sampler, and therefore to slower convergence.}

@@

It is also possible, especially during the early stages of the
algorithm, where the mixture proposal is a poor approximation of the
target distribution, for a sample to be produced with a weight which
is orders of magnitude bigger than the rest. This is a problem, since
the resampling step will then lead to a vector of particles all
centred at the outlier sample. Fortunately these spikes in weights can
easily be detected, and in such a case, the sample for this iteration
can be removed. The ETPF can then be used to formulate a new mixture
sample with one particle placed at the outlier point, and the others
distributed according to the particle positions before the problematic
iteration. This ensures that the region where the problem occurred is
better represented by the mixture, and prevents further spikes in this region.

@@

The statistic $\neff$ is easier to deal with than the variance of the
weights, as it takes values on $[1, M]$ while the variance of the
weights takes values on $\mathbb{R}_\geq$. Moreover the variance of
the weights can vary
over many orders of magnitude causing numerical instabilities, so that the effective
sample size is more desirable as an indicator of optimality.

@@

The effective sample size also has another useful property; if we
consider the algorithm in the early stages, for example, we have all
$M$ particles in the tails of the target searching for a mode. The
particle closest to the mode will have an exponentially higher weight
and the effective sample size will be close to 1. In later stages the
ensemble populates high density regions, and better represents the
posterior distribution. This leads to smaller variation in the
weights, and a higher effective sample size. By looking for
approximations of the effective sample size which look like a
stationary distribution, we can tell when the algorithm is working
efficiently.

@@

\section{Approximate Multinomial Resampling}\label{sec:MT}

@@

\begin{table}[!ht]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	$\b{z} = M\bar{\b{w}}$.\;
	\For {$i = 1,\dots, M$}
	{
		$J = \argmax_j z_j$.\;
		$p_{i,J} = \min\{1,z_J\}$.\;
		$z_J = z_J - p_{i,J}$.\;
		\While {$\sum_j p_{i,j} <1$}
		{
			$K = \argmin_{k \in \{k|z_k>0\}} \|y_J - y_k\|$.\;
			$p_{i,K} = \min\{1-\sum_j p_{i,j}, z_K\}$.\;
			$z_K = z_K - p_{i,K}$.\;
		}
		$x_i = \sum_k p_{i,k}y_k$.\;
	}
\caption{The approximate multinomial resampler (MT).\label{alg:MT}}
\end{algorithm}
\end{table}

@@

Case two is slightly more involved. Estimation of the normalisation constant $Z$ is biased, and so estimates of statistics are sums of independent but biased estimators. Since the estimators are independent, the proof of consistency of ETAIS in this second limit can be approached in the same way as the PMC algorithm, where it has been shown that $\hat{Z}\rightarrow Z$~\cite{robert2013monte}. Since the normalisation constant is consistent, sums of the independent estimators are also consistent.

@@

{\red
\section{A useful property of the ETAIS algorithm for multimodal
  distributions} } \label{sec:useful}

@@

{\red The numerics contained herein compare the ETAIS approach that we
  have presented against Metropolis-Hastings methods with
  the same type of proposal as is used by the kernels within
  ETAIS. Since we propose that ETAIS is a potential framework for
  parallelised Bayesian computations, we compare ETAIS with $N$
  ensemble members against an independent ensemble of $N$
  Metropolis-Hastings chains. All implementations were
  computed in serial, but the results demonstrate a clear speed-up
  which could be further exploited through parallelisation in the future.
  }

@@

MH methods can struggle
with multimodal problems, particularly where switches between the
modes are rare, resulting in incorrectly proportioned modes in the
histograms. This example demonstrates that the ETAIS algorithm
redistributes particles to new modes as they are found. This means
that we expect the number of particles in a mode to be approximately
proportional to the probability density in that mode, resulting in
faster global convergence. {\red In particular, we will look at the effect
of using either the ETPF, MT or a standard bootstrap resampler.

@@

Figure \ref{kshdbf} shows the bimodal target density, and convergence
plots for the RWMH and ETAIS-RW with ETPF, MT and bootstrap
resamplers. As expected, a higher quality resampler leads to better
proposal distributions, which in turn leads to greater stability and
more reliable convergence. However, this also shows that the MT is a
good greedy approximation of the ETPF, and for a fraction of the cost
when the number of ensemble members is larger, as shown in Section
\ref{sec:MT}. The RWMH algorithm fails to converge efficiently since chains only very rarely
make switches between the two modes, leading to very slow mixing. This
demonstrates the advantage that ensemble-based methods can have over
other methods when the target
is multimodal.

@@

For this example, we observe noisily the position of a particle at ten
equally spaced points in the time interval $t \in (0, 1]$. The Lorenz
equations are solved with the chaotic parameters given above and the
initial condition $\mathbf{x}_0^1$. A time step of $h=1\times 10^{-3}$
was used to evolve the equations numerically with the explicit Euler
method. The noise added to each observation was taken from the
distribution {\red $\mathcal{N}(0, 0.1^2I)$.} Priors for the initial condition
coordinates were given by
\[
	x_0 \sim \mathcal{N}(-0.5, 0.03^2), \quad y_0 \sim\mathcal{N}(-0.5, 0.03^2) \quad \text{and} \quad z_0 \sim \mathcal{N}(15, 0.4^2).
\]
The posterior density function takes the form
\[
	\pi(\mathbf{x}_0|D) \propto \exp\left\{-\frac{1}{2}\|\mathcal{G}(\mathbf{x}_0) - D\|^2_\Sigma\right\}\mathcal{N}(\mathbf{x}_0; m, 0.03^2\text{I}),
\]
where $\mathcal{G}$ calculates the solution to the Lorenz equations, using the Euler method with a time step of $h=1\times 10^{-3}$ starting at the initial condition $\mathbf{x}_0$, and returns the position at $t = 0.1, ..., 1$. The mean of the prior distribution $m = [-0.5, -0.5, 15]$.

@@

{\red The implementations of the ETAIS that have been presented in this
  paper, have been run in serial, but we argue that, particularly for
  inverse problems with expensive likelihoods, for example in the case
  where this requires the approximation of the solution of a
  differential equation, that this method is an excellent candidate
  for parallelisation, with the likelihood evaluations for each
  particle being computed on a different processor. This said, even in
  parallel, the numerical results that we have presented demonstrate
  that even a serial implementation can outperform standard
  Metropolis-Hastings methods.}

>>>>

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Ensemble Transport Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. {\red As applications of these methods increase in size and
  complexity, the need for efficient methods increases. In this
paper, we present a particle ensemble algorithm.} At each iteration, an
  importance sampling proposal distribution is formed using an
  ensemble of particles. A stratified sample is taken from this
  distribution and weighted under the posterior, a state-of-the-art
  ensemble transport resampling method is then used to create an evenly weighted sample
  ready for the next iteration. We demonstrate that this ensemble transport
  adaptive importance sampling (ETAIS) method outperforms MCMC methods
  {\red with equivalent proposal distributions} for low dimensional problems,
  and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, {\red multinomial transformation (MT),} which while not
  as accurate as the ensemble transport resampler, is substantially less costly for large
  ensemble sizes, and can then be used in conjunction with ETAIS for
  complex problems. We also focus on how algorithmic parameters
  regarding the mixture proposal can be quickly tuned to optimise
  performance. In particular, we demonstrate this methodology's
  superior sampling for multimodal problems, such as those arising
  from inference for mixture models, and for problems with expensive
  likelihoods requiring the solution of a differential equation, {\red
    for
  which speed-ups of orders of magnitude are demonstrated. Likelihood
  evaluations of the ensemble could be computed in a distributed
  manner, suggesting that this methodology is a good candidate
  for parallel Bayesian computations.}
\end{abstract}
\begin{keywords}MCMC, importance sampling, Bayesian, inverse problems,
  {\red ensemble, resampling}.
\end{keywords}
\section{Introduction}
Having first been developed in the early 1970s\cite{hastings1970monte}, Markov chain Monte Carlo (MCMC) methods have been of increasing
importance and interest in the last 20 years or so. They allow us to
sample from complex probability distributions which we would not be
able to sample from directly. In particular, these methods have
revolutionised the way in which inverse problems can be tackled,
allowing full posterior sampling when using a Bayesian framework.

@@

Then, according to Bayes' theorem, the posterior density that we
are interested in characterising is given, up to a constant of proportionality, by
\begin{equation*}
	\pi \propto \exp \left ( -\Phi(x^*) \right )\pi_0(x^*),
\end{equation*}
where $\pi_0$ is the prior density on the quantity of interest $x^*$.

@@

\subsection{Particle filters and resamplers}\label{sec:filters}
In this subsection we briefly review particle filters, since the
development of the resampler that we incorporate into the ETAIS is
motivated by this area. Particle filters are a class of Monte Carlo algorithm designed to
solve the filtering problem. That is, to find the best estimate of the
true state of a system when given only noisy observations of the
system. The solution of this problem has been of importance since the
middle of the 20th century in fields such as molecular biology,
computational physics and signal processing. In recent years the data
assimilation community has contributed several efficient particle
filters, including the ensemble Kalman filter
(EnKF)~\cite{evensen1994sequential} and the ensemble transform
particle filter (ETPF)~\cite{reich2013nonparametric}.

@@

In this work we use the resampler used within these particle filters. We
approximate the posterior {\red density $\pi$, with an ensemble
of weighted samples,} $\{(w_i,y_i)\}_{i=1}^M$. In filtering problems
the weights would be found by incorporating new observed data whereas
here we simply use importance weights. These weighted samples can then be resampled, using the ETPF or otherwise,
into a {\red set of new equally-weighted samples $\{x_i\}_{i=1}^M$, such that
\begin{equation}\label{eqn:resampler}
	\sum\limits_{i=1}^M \! w_i\delta_{y_i}(\cdot)
	 \quad \xrightarrow{\text{ETPF}} \quad \sum\limits_{i=1}^M \!
	 \frac{1}{M} \delta_{x_i}(\cdot),
\end{equation}}
{\red where $\delta_{x}(\cdot)$ is the Dirac delta measure.}

The resamplers used within particle filters such as the ETPF are well suited to this problem
since it is easy to introduce conditions in the resampling to ensure
you obtain the behaviour you require. One downside is that the
required ensemble size increases quickly with dimension, making it
difficult to use in high-dimensional problems.

@@

Initially we
will use the ETPF resampler algorithm\cite{reich2013nonparametric}, although we
will suggest an alternative strategy in Section \ref{sec:MT}, for
examples where a
large ensemble is required, and for which the ETPF may become more expensive. The
resampling algorithm gives us a set of evenly weighted samples which
represents the target distribution well, which we can use to iterate
the process again. The algorithm is summarised in Algorithm
\ref{alg:ETAIS}.

{\red The choice of resampling algorithm is important since its
  output is used to formulate the proposal distribution for the next
  iteration. A basic resampler may be cheap to implement, but the
  resulting sample, and in turn the next iteration's proposal distribution, may not be as representative of the target
  distribution. This leads to higher variances of the weights in the
  importance sampler, and therefore to slower convergence to the
  target density.}

@@

It is also possible, especially during the early stages of the
algorithm, where the mixture proposal is a poor approximation of the
target distribution, for a sample to be produced with a weight which
is orders of magnitude bigger than the rest. This is a problem, since
the resampling step will then lead to a vector of samples all
centred at the outlier sample. Fortunately these spikes in weights can
easily be detected, and in such a case, the sample for this iteration
can be removed. The ETPF can then be used to formulate a new mixture
sample with one sample placed at the outlier point, and the others
distributed according to the sample positions before the problematic
iteration. This ensures that the region where the problem occurred is
better represented by the mixture, and prevents further spikes in this region.

@@

The statistic $\neff$ is easier to deal with than the variance of the
weights, as it takes values on $[1, M]$ while the variance of the
weights takes values on $\mathbb{R}^+$. Moreover the variance of
the weights can vary
over many orders of magnitude causing numerical instabilities, so that the effective
sample size is more desirable as an indicator of optimality.

@@

The effective sample size also has another useful property; if we
consider the algorithm in the early stages, for example, we have all
$M$ samples in the tails of the target searching for a mode. The
sample closest to the mode will have an exponentially higher weight
and the effective sample size will be close to 1. In later stages the
ensemble populates high density regions, and better represents the
posterior distribution. This leads to smaller variation in the
weights, and a higher effective sample size. By looking for
approximations of the effective sample size which look like a
stationary distribution, we can tell when the algorithm is working
efficiently.

@@

\section{Multinomial Transformation}\label{sec:MT}

@@

\begin{table}[!ht]
\centering
\begin{algorithm}[H]
\DontPrintSemicolon
\BlankLine
	$\b{z} = M\bar{\b{w}}$.\;
	\For {$i = 1,\dots, M$}
	{
		$J = \argmax_j z_j$.\;
		$p_{i,J} = \min\{1,z_J\}$.\;
		$z_J = z_J - p_{i,J}$.\;
		\While {$\sum_j p_{i,j} <1$}
		{
			$K = \argmin_{k \in \{k|z_k>0\}} \|y_J - y_k\|$.\;
			$p_{i,K} = \min\{1-\sum_j p_{i,j}, z_K\}$.\;
			$z_K = z_K - p_{i,K}$.\;
		}
		$x_i = \sum_k p_{i,k}y_k$.\;
	}
\caption{The  multinomial transformation (MT).\label{alg:MT}}
\end{algorithm}
\end{table}

@@

Case two is slightly more involved. Estimation of the normalisation constant $Z$ is biased, and so estimates of statistics are sums of independent but biased estimators. Since the estimators are independent, the proof of consistency of ETAIS in this second limit can be approached in the same way as the pMC algorithm, where it has been shown that $\hat{Z}\rightarrow Z$~\cite{robert2013monte}. Since the normalisation constant is consistent, sums of the independent estimators are also consistent.

@@

\section{{\red A useful property of the ETAIS algorithm for multimodal
  distributions } }\label{sec:useful}

@@

{\red The numerics contained herein compare the ETAIS approach that we
  have presented against Metropolis-Hastings methods with
  the same type of proposal as is used by the kernels within
  ETAIS. Since we propose that ETAIS is a potential framework for
  parallelised Bayesian computations, we compare ETAIS with $N$
  ensemble members against an independent ensemble of $N$
  Metropolis-Hastings chains. All implementations were
  computed in serial, but the results demonstrate a clear speed-up
  which could be further exploited through parallelisation.
  }

@@

MH methods can struggle
with multimodal problems, particularly where switches between the
modes are rare, resulting in incorrectly proportioned modes in the
histograms. This example demonstrates that the ETAIS algorithm
redistributes samples to new modes as they are found. This means
that we expect the number of samples in a mode to be approximately
proportional to the probability density in that mode, resulting in
faster global convergence. {\red In particular, we will look at the effect
of using either the ETPF, MT or a standard bootstrap resampler.

@@

Figure \ref{kshdbf} shows the bimodal target density, and convergence
plots for the RWMH and ETAIS-RW with ETPF, MT and bootstrap
resamplers, averaged over 32 repeats. As expected, a higher quality resampler leads to better
proposal distributions, which in turn leads to greater stability and
more reliable convergence. However, this also shows that the MT is a
good greedy approximation of the ETPF, and for a fraction of the cost
when the number of ensemble members is larger, as shown in Section
\ref{sec:MT}. The RWMH algorithm fails to converge efficiently since chains only very rarely
make switches between the two modes, leading to very slow mixing. This
demonstrates the advantage that ensemble-based methods can have over
other methods when the target
is multimodal.

@@

For this example, we observe noisily the position of a particle at ten
equally spaced points in the time interval $t \in (0, 1]$. The Lorenz
equations are solved with the chaotic parameters given above and the
initial condition $\mathbf{x}_0^1$. A time step of $h=1\times 10^{-3}$
was used to evolve the equations numerically with the explicit Euler
method. The noise added to each observation was taken from the
distribution {\red $\mathcal{N}(0, 0.1^2I)$.} Priors for the initial condition
coordinates were given by
\[
	x_0 \sim \mathcal{N}(-0.5, 0.4^2), \quad y_0 \sim\mathcal{N}(-0.5, 0.4^2) \quad \text{and} \quad z_0 \sim \mathcal{N}(15, 0.4^2).
\]
The posterior density function takes the form
\[
	\pi(\mathbf{x}_0|D) \propto \exp\left\{-\frac{1}{2}\|\mathcal{G}(\mathbf{x}_0) - D\|^2_\Sigma\right\}\mathcal{N}(\mathbf{x}_0; m, 0.4^2\text{I}),
\]
where $\mathcal{G}$ calculates the solution to the Lorenz equations, using the Euler method with a time step of $h=1\times 10^{-3}$ starting at the initial condition $\mathbf{x}_0$, and returns the position at $t = 0.1, ..., 1$. The mean of the prior distribution $m = [-0.5, -0.5, 15]$.

@@

{\red The implementations of the ETAIS that have been presented in this
  paper, have been run in serial, but we argue that, particularly for
  inverse problems with expensive likelihoods, for example in the case
  where this requires the approximation of the solution of a
  differential equation, that this method is an excellent candidate
  for parallelisation, with the likelihood evaluations for each
  particle being computed on a different processor. This said, the numerical results that we have presented demonstrate
  that even a serial implementation can outperform standard
  Metropolis-Hastings methods.}
